MaBoSS for HPC environments: Implementations of the continuous time Boolean model simulator for large CPU clusters and GPU accelerators

### 2024-05-02 Article.tex
connect output section with cpu section

MaBoSS was initially developed as a single-core application, but very quickly it was extended with a basic parallelism to exploit the multi-core nature of modern CPUs. In this parallel implementation the simulation of trajectories, as well as the statistics aggregation, was distributed on multiple cores (controlled by a simulation parameter) using POSIX threads. In the following sections of the papers, we will refer to this implementation as the CPU version.

@@

Each of the three kinds of statistics is in its nature a sample from a probabilistic distribution of Boolean states. This sample is represented in code as a hash map in the CPU version, varying in the $(key, value)$ pairs according to the specific statistic. For the final states, the keys of the hash map are the states of the model, and the values are the number of times the state was sampled last in a trajectory. Such output can be visualized as a pie chart (see Figure~\ref{fig:cohen_final}). The fixed states are represented similarly, but only the fixed points are stored in the hash map as keys.

@@

The final and fixed state statistics characterize the behavior of the model at one point in time --- at the end of the simulation. The network state probabilities on a time window highlight more dynamic characteristics of the model, showing how the average trajectory evolves over the simulation time. Programmatically, it is an extension of the final state statistics but instead of one hash map, there is a hash map for each time window. The hash map values are the state durations in the specific time window aggregated over all simulated trajectories. Further, these statistics can be visualized in various ways using a line chart. Figure~\ref{fig:cohen_traj} shows which non-internal nodes are active throughout the simulation.

As mentioned at the beginning of the section, if the trajectory does not reach a fixed point, the simulation is stopped after the maximal allowed time. This is a common scenario especially when some trajectories form circles, i.e. when a model has \emph{circular steady states}. A circular steady state is usually not directly visible from the state probability line charts; Stoll et al.~\cite{stoll2012continuous} proposed methods to detect them (such as plotting the state and transition entropies) but we do not discuss them further in this paper for the sake of brevity.

>>>>

MaBoSS was initially developed as a single-core application, but swiftly, it was extended with a basic parallelism to exploit the multi-core nature of modern CPUs. In this parallel implementation, the simulation of trajectories and the statistics aggregation were distributed among multiple cores using POSIX threads. In the following sections of the papers, this implementation will serve as a baseline, and we will refer to it simply as the \emph{CPU version}.

Each statistics data held by a thread is represented by a hash map with the keys as the states of the model and the values as a numerical value. Therefore, their aggregation from multiple trajectories of multiple threads is carried out by a well-researched parallel sum reduction. To better understand how a researcher can use MaBoSS output, in the following section, we discuss the differences between the statistics in greater detail and show the standard ways of their visualization.

@@

Each of the three kinds of statistics is in its nature a sample from a probabilistic distribution of Boolean states. This sample is represented in code as a hash map in the CPU version, varying in the $(key, value)$ pairs according to the specific statistic. For the final states, the keys of the hash map are the model states, and the values are the number of times the state was sampled as the last in a trajectory. Such output can be visualized as a pie chart (see Figure~\ref{fig:cohen_final}). The fixed states are represented similarly, but only the fixed points are stored in the hash map as keys.

@@

The final and fixed state statistics characterize the behavior of the model at one point in time --- at the end of the simulation. The network state probabilities on a time window highlight more dynamic characteristics of the model, showing how the average trajectory evolves over the simulation time. Programmatically, it is an extension of the final state statistics --- instead of one hash map, there is a hash map for each time window. The hash map values are the state durations in the specific time window aggregated over all simulated trajectories. Further, these statistics can be visualized in various ways using a line chart. Figure~\ref{fig:cohen_traj} shows which non-internal nodes are active throughout the simulation.

As mentioned at the beginning of the section, if the trajectory does not reach a fixed point, the simulation is stopped after the maximal allowed time. This is a common scenario, especially when some trajectories form circles, i.e., when a model has \emph{circular steady states}. A circular steady state is usually not directly visible from the state probability line charts; Stoll et al.~\cite{stoll2012continuous} proposed methods to detect them (such as plotting the state and transition entropies), but we do not discuss the methods further in this paper for the sake of brevity.
